# Bus passenger changes

Unlike traffic counter data, bus patronage data is a direct measure of people- it is people boarding buses in the network and tagging on (including transfers). 

```{r}
# -----------------------------
# 1. Libraries that are needed 
# -----------------------------

library(dplyr)
library(readr)
library(lubridate)
library(slider)
library(sf) # handling spaital data
library(ggplot2) # making the map
library(ggrepel) # if you want labels with lines
source("../support_files/theme.R")
source("../support_files/core_functions.R")

# -----------------------------
# Bus data
# -----------------------------

basic_readings = bus_read_data()
subdivided = subdiv_before_after(basic_readings)

```

```{r}
#| fig-cap: "For bus routes present in 2025 and 2026, change in March-May bus passengers compared to expectations based on 2025"
grf1 <- subdivided |> 
  mutate(to_nearest = round(delta_percent,0)) |> 
  #ggplot(aes(x = to_nearest)) +
  ggplot(aes(x = delta_percent)) +
  geom_histogram(binwidth = 5,fill = five_cols[3]) +
  geom_vline(xintercept = 0, linewidth=0.05)+
  #geom_dotplot(method = "histodot", binaxis = "x", stackdir = "up", binwidth=1, dotsize = 1, fill = five_cols[3]) +
  theme_david() +
  theme(axis.text.y.left = element_blank(),
        axis.line.y.left = element_blank(),
        axis.ticks.y.left = element_blank(),
        axis.title.y.left= element_blank()) +
  labs(x="Percentage change v. expected use")
figuresave(x=grf1, fname="49_bus_routes")
grf1
```
I haven't had time to unpick individual changes to routes that might have effected route specific use, but we can broadly say that the changes in the individual routes is consistent with a spread around the 11% higher use of the total network. This to me suggests it is a general change.


```{r}
#| fig-cap: "Rolling 7 day total daily bus patronage"
#trend
trendata <- rolling7_total(basic_readings)

rollgraph <- rolling_graph(trendata)
figuresave(x=rollgraph, fname="50_bus_rolling")
rollgraph
```
The year began with passenger boarding numbers notably down on last year, associated with trade-offs in changes in fare structure. About a week after the outbreak of war, passenger numbers rose. This rise to levels equal to or above last year has remained for the full period.

```{r}
#| fig-cap: "Weekday percentage change in bus boardings compared to 2025 expectations"
## Weekday changes
weekDayed <- aggregate_by_weekday(basic_readings)
wkplt <- weekplot(weekDayed)
figuresave(x=wkplt, fname="51_bus_weekdays")
wkplt
```
Bus changes are at their lowest on Wednesday and Monday, which mirrors both Light and Heavy Vehicles. Other weekdays have notably rises.
