# Cycle count changes

As counter data, the cycle count data is similar to vehicle count data. It has similar issues around road works etc affecting individual counter results.

```{r}
# -----------------------------
# 1. Libraries etc
# -----------------------------

library(dplyr)
library(readr)
library(lubridate)
library(slider)
library(sf) # handling spaital data
library(ggplot2) # making the map
library(ggrepel) # if you want labels with lines
source("../support_files/theme.R")
source("../support_files/core_functions.R")

# -----------------------------
# topmap
# -----------------------------

cycle_readings = cycle_read_data()
basemap = create_basemap()

# prepping the top layer
subdivided = subdiv_before_after(cycle_readings)
cycle_latlong <- read_csv("../Data/DCC/counternames.csv") |> 
  filter(type=="cycle")
sf_points <- st_as_sf(cycle_latlong, coords = c("long", "lat"), crs = 4326)
sf_nztm <- st_transform(sf_points, crs = 2193)
coords <- st_coordinates(sf_nztm)
cycle_latlong$X <- coords[,1]
cycle_latlong$Y <- coords[,2]

placed <- cycle_latlong |> 
  rename(subdivision = countloc) |> 
  inner_join(subdivided, by = join_by(subdivision)) 
```

```{r}
#| fig-cap: "For counters available in 2025 and 2026, 2026 change in March-May cycle traffic counts compared to expectations based on 2025"
# make a map
m1 <- basemap +
  geom_point(data=placed, aes(x=X,y=Y), size=0.5, colour="red") +
  geom_label_repel(data=placed, aes(x=X,y=Y, label=delta_percent_format),
                   max.overlaps = Inf, size=4, fill="#FFFFFF99",
                   point.padding = 0.1, box.padding = .5,
                   min.segment.length = 0) 
figuresave(x=m1, fname="52_cycle_counters")
m1
```
Many of the individual cycle counters around town are up from between 15% and 25%. Another set of peripheral sites are in the 0-9% range. A few outliers have other readings.


```{r}
#| fig-cap: "Rolling 7 day total daily cycle counts"
#trend
trendata <- rolling7_total(cycle_readings)
rollgraph <- rolling_graph(trendata)
figuresave(x=rollgraph, fname="53_cycle_rolling")
rollgraph
```
The year began with noticeably lower total daily cycle counts, which (like bussing) a week after the war outbreak rose to levels the same or above 2025 based expectations.

```{r}
#| fig-cap: "Weekday percentage change in cycle counts compared to 2025 expectations"
## Weekday changes
weekDayed <- aggregate_by_weekday(cycle_readings)
wkplt <- weekplot(weekDayed)
figuresave(x=wkplt, fname="54_cycle_weekdays")
wkplt

```
As with vehicles or buses, cyclists are lower on a Wednesday, and Monday has the second lowest growth compared to expectations. This does seem to be a general trend in degree of movement, rather than just the means of movement. 
