# Overall comments

```{r}
library(ggplot2)
source("../support_files/core_functions.R")
source("../support_files/theme.R")
```


```{r}
bus_starter <- bus_read_data()
bus_pre_post <- before_after(bus_starter)
cycle_starter <- cycle_read_data()
cycle_pre_post <- before_after(cycle_starter)
pedestrian_starter <- pedestrian_read_data()
pedestrian_pre_post <- before_after(pedestrian_starter)
light_starter <- NZTA_read_data("Light")
light_pre_post <- before_after(light_starter)
heavy_starter <- NZTA_read_data("Heavy")
heavy_pre_post <- before_after(heavy_starter)
```

```{r}
#| fig-cap: "All modes of transport. Change in daily average counts in March-May 2026 compared to expectations based on 2025"
#| fig-height: 3
gfdf <- bind_rows(bus_pre_post, cycle_pre_post, heavy_pre_post,
          light_pre_post, pedestrian_pre_post) |> 
  filter(Yr == "2026") |> 
  mutate(x1 = if_else(Mn == "pre-crisis", ymd("2026-01-01"), ymd("2026-03-01")),
         x2 = if_else(Mn == "pre-crisis", ymd("2026-02-28"), ymd("2026-05-31")))

gfit <- allgraph(gfdf)
figuresave(x=gfit, fname="58_all")
gfit
```

There is a general mode shift away from cars and towards other transport. As there is a baseline of more car use, small percentage drops in personal vehicle use are associate with large percentage increases of their baselines in other modes of transport.

But it is not just mode shift. Checking the days of the week suggests less travel across all modes on (particularly) Wednesday and (less so) Monday. This does look a lot like what you would see if there was more work from home taking place.