---
title: "War, fuel, and transport changes"
author: "David Hood, david@thoughtful.net.nz"
date: today
abstract: "An analysis of changes in transport use since 28th February 2026 for Ōtepoti/Dunedin. Across all sources, there is strong evidence that there has been a shift in modes."

format:
  pdf:
    pdf-engine: xelatex
    include-in-header: .\preamble.tex
    toc: false
    toc-depth: 2
    number-sections: true
    fig-pos: H
    fig-cap-location: bottom
    fig-device: ragg_png
    dpi: 300

# Document layout
fontsize: 11pt
geometry: margin=1in
linestretch: 1.3

# Figures (global defaults)
fig-width: 5
fig-height: 3
fig-align: center
fig-format: png

# Code execution
execute:
  echo: false
  warning: false
  message: false

# Optional metadata you can reuse later
keywords:
  - transport
  - fuel
  - analysis
---

```{r}
library(ggplot2)
source("../support_files/core_functions.R")
```


```{r all_the_data}
bus_starter <- bus_read_data()
bus_pre_post <- before_after(bus_starter)
cycle_starter <- cycle_read_data()
cycle_pre_post <- before_after(cycle_starter)
pedestrian_starter <- pedestrian_read_data()
pedestrian_pre_post <- before_after(pedestrian_starter)
light_starter <- NZTA_read_data("Light")
light_pre_post <- before_after(light_starter)
heavy_starter <- NZTA_read_data("Heavy")
heavy_pre_post <- before_after(heavy_starter)
```
\vspace*{1.0\baselineskip}
\begin{execsummary}
Compared to the same time last year, among traffic evidence for Ōtepoti/Dunedin:

- Fuel is up (Regular Petrol +28\%, Diesel +74\%)

- Vehicle traffic counts are down (Light -2.6\%, Heavy -1.1\%)

- Bus patronage is up 11\%

- Cycle traffic counts are up 14.4\%

- Pedestrian traffic counts are up 34.3\%

\end{execsummary}

```{r}
#| fig-height: 3
gfdf <- bind_rows(bus_pre_post, cycle_pre_post, heavy_pre_post,
          light_pre_post, pedestrian_pre_post) |> 
  filter(Yr == "2026") |> 
  mutate(x1 = if_else(Mn == "pre-crisis", ymd("2026-01-01"), ymd("2026-03-01")),
         x2 = if_else(Mn == "pre-crisis", ymd("2026-02-28"), ymd("2026-05-31")))

gfit <- allgraph(gfdf)
figuresave(x=gfit, fname="00_all")
gfit
```

\newpage

\tableofcontents



\pagebreak
{{< include ../chapters/01_introduction.qmd >}}

\pagebreak
{{< include ../chapters/02_fuel_prices.qmd >}}

\pagebreak
{{< include ../chapters/03_census.qmd >}}

\pagebreak
{{< include ../chapters/04_light.qmd >}}

\pagebreak
{{< include ../chapters/05_heavy.qmd >}}

\pagebreak
{{< include ../chapters/06_bus.qmd >}}

\pagebreak
{{< include ../chapters/07_cycle.qmd >}}

\pagebreak
{{< include ../chapters/08_pedestrian.qmd >}}

\pagebreak
{{< include ../chapters/09_mode_shift.qmd >}}
